### 2023-02-24 manuscript.Rmd
update author contrib
Add discussion paragraph, minor edits

author: 
  - name          : "Elisa S. Buchberger"
    affiliation   : "1"
    corresponding : yes
    address       : "Lentzeallee 94, 14195 Berlin"
    email         : "buchberger@mpib-berlin.mpg.de"
  - name          : "Chi T. Ngo"
    affiliation   : "1"
  - name          : "Aaron Peikert"
    affiliation   : "1, 2"
  - name          : "Andreas M. Brandmaier"
    affiliation   : "1,2,3"
  - name          : "Markus Werkle-Bergner"
    affiliation   : "1"

@@

authornote: |
  This work was conducted within the project “Lifespan Rhythms of Memory and Cognition (RHYME)” at the Max Planck Institute for Human Development. ESB is a fellow of the International Max Planck Research School on the Life Course (LIFE; http://www.imprs-life.mpg.de/en). AP is a fellow of the International Max Planck Research School on Computational Methods in Psychiatry and Ageing Research (COMP2PSYCH, https://www.mps-ucl-centre.mpg.de/comp2psych). MW-B received support from the Jacobs Foundation (Early Career Research Fellowship 2017–2019). CTN’s work is supported by a grant from the German Research Foundation (DFG; NG 191/2-1) and an Early Career Research Fellowship by the Jacobs Foundation (2021-1417-99). We are grateful to all members of the RHYME and LIME projects for valuable feedback on the project. Finally, we acknowledge support by the Max Planck Dahlem Campus of Cognition (MPDCC).
  The latest supplementary materials for this article are available at <https://github.com/ebuchberger/Estimating-Power-for-SEM>. The version of code and data used as the basis for this PDF is "`r paste0(repro::current_hash(backend = 'gert'))`".

@@

# 2. Methods
In the following, we demonstrate how to conduct a simulation-based power analysis for non-nested model comparisons. For the sake of illustration, we will refer to the example on the compositional structure of memory outlined in section 1.3 throughout. For this example, we aim to derive estimates of required sample size to identify the tri-partite model as the best fitting model, if it indeed generated the data. In a step-by-step guide, we therefore show how to specify the competing models in R, how to set up different design conditions for the simulation, how to define the functions to generate, analyze and summarize synthetic data and how to execute the simulation. Based on two separate simulations, we show how the separability of theoretical constructs (section 3.1) and the reliability of the manifest variables (section 3.2) impacts estimates of statistical power for model comparisons by varying inter-factor correlations and factor loadings across different conditions of the simulation. Finally, we employ the same simulation-based approach to investigate Type-I error rates, that is erroneously identifying the tri-partite model as the best fitting model if the bi-partite or unitary model generated the data (section 3.3).

@@

The model is defined by a model string, which specifies three latent factors of memory (Pattern Separation [ps]; Pattern Completion, [pc]; Generalization, [gen]) which are measured through the manifest variables (asks 1 to 13). Note that lavaan entails several user-friendly functions for fitting models that automatically include default settings for model components that are not explicitly specified (e.g. adding residual variances, covariances of exogenous latent variables). Thus, the model syntax can be kept very concise (check documentation of the respective function for full information on default settings).

@@

## 3.2	What is the Impact of the Reliability of Manifest Variables on Estimates of Statistical Power?

In a second step, we further examined the effect of the construct reliability on statistical power by investigating the loading strength of the manifest variables. To this end we conducted a second simulation following the steps outlined above while varying the sample size from 50 to 200 and the loadings of the manifest variables from 0.5 to 0.9. Figure 3B displays the obtained power curves, illustrating statistical power dependent on total sample size and loading strength of the manifest variables. For this estimate, the factor inter-correlation was set to 0.3. Again, increasing sample size lead to higher statistical power. However, the loading strength impacted estimates of statistical power: the higher the loadings of the manifest variables, the higher the statistical power achieved by a given sample size. For reasonably high levels of factor loadings (>= 0.7) a statistical power of 95% was reached for *N* = 125.

@@

First, we showed that detecting differences in model fit requires increasingly many participants the higher the inter-correlation of the factors turns out practically. Therefore, it is warranted to consider the actual separability of the constructs in question to avoid underpowered sampling plans. We acknowledge that committing to specific parameter values, such as factor inter-correlations, can be difficult when setting up the population models for the simulations. Thus, we recommend researchers to consult existing literature in order to make informed assumptions on the relationship among different latent factors. For cases in which no or little evidence exists on the relationship between the constructs under investigation, we recommend to enter these inter-correlations into the simulation as parameter of interest (as demonstrated in the example of this paper) and to then carefully evaluate what level of resolution is required when separating inter-related factors depending on practical relevance.  
Second, the results on the loadings of manifest variables stresses the importance of employing indicators that capture the latent construct with high reliability [@zuo_2019]. Considering the issue of reliability is a crucial step in experimental design, as the reliability of the indicators heavily influences the practical ability to identify the targeted latent construct with sufficient precision and differentiate between competing models. Again, committing to population values can be difficult, especially if there is no prior work from which to draw values. In this case, we still advice to estimate plausible values from related scientific areas. For instance, selecting factor loading strength in the tri-partite model is not straightforward. However, we can do a literature search on similar models and take the reported test- reliability (e.g., a test-retest) or construct reliability (e.g., coefficient alpha) and translate this back into factor loadings strength and residual loadings [see @brandmaier_2018 for examples on how to parametrize longitudinal models from minimal available information]. If information about test reliability is available, we set the loadings to the square-root of the reliability to take this prior information into account. Then, this yields the following construct reliability for our design (assuming identical loadings across *k* items):

@@

While we demonstrate in this tutorial that simulation-based techniques help overcome methodological shortcomings from traditional analytical approaches, the randomness that is inherent to such approaches may seem at odds with efforts in results reproducibility and replicability. Here, a clear-cut solution is through transparent and accessible documentation of the simulation code and adequate software management. That is, researchers should make use of repositories on platforms such as GitHub or the Open Science Framework to make code scripts available to other researchers, and clearly specify the versions of software programs and packages that were used (e.g. via containerization or at least in written form). We thus urge researchers to seriously consider the issue of reproducibility for their simulation analyses, to make outcomes of such a-priori power estimates easily accessible to other researchers and reviewers [for a more detailed guideline on a reproducible workflow in R see @peikert_2021 and the implementation of this manuscript on GitHub as an example]. Note, however, that—specifically in cases when simulations are parallelized on computing clusters—the exact reproducibility of simulations as demonstrated in this tutorial might be compromised, even when a seed is specified in the analysis script. Nevertheless, a sufficiently high number of iterations ensures replicability, i.e., comparible estimates of statistical power. Derived conclusions about required sample size will therefore remain unaffected.

@@

>>>>

author: 
  - name          : "Elisa S. Buchberger"
    affiliation   : "1"
    corresponding : yes
    address       : "Lentzeallee 94, 14195 Berlin"
    email         : "buchberger@mpib-berlin.mpg.de"
  - name          : "Chi T. Ngo"
    affiliation   : "1"
  - name          : "Aaron Peikert"
    affiliation   : "1, 2"
  - name          : "Andreas M. Brandmaier"
    affiliation   : "1,2,3"
    equal_contrib: yes
  - name          : "Markus Werkle-Bergner"
    affiliation   : "1"
    equal_contrib : yes

@@

authornote: |
  This work was conducted within the project “Lifespan Rhythms of Memory and Cognition (RHYME)” at the Max Planck Institute for Human Development. ESB is a fellow of the International Max Planck Research School on the Life Course (LIFE; http://www.imprs-life.mpg.de/en). AP is a fellow of the International Max Planck Research School on Computational Methods in Psychiatry and Ageing Research (COMP2PSYCH, https://www.mps-ucl-centre.mpg.de/comp2psych). MW-B received support from the Jacobs Foundation (Early Career Research Fellowship 2017–2019, 2016-1217-16). CTN’s work is supported by a grant from the German Research Foundation (DFG; NG 191/2-1) and an Early Career Research Fellowship by the Jacobs Foundation (2021-1417-99). We are grateful to all members of the RHYME and LIME projects for valuable feedback on the project. Finally, we acknowledge support by the Max Planck Dahlem Campus of Cognition (MPDCC). The authors declare no competing interests.
  The latest supplementary materials for this article are available at <https://github.com/ebuchberger/Estimating-Power-for-SEM>. The version of code and data used as the basis for this PDF is "`r paste0(repro::current_hash(backend = 'gert'))`".

@@

# 2. Methods
In the following, we demonstrate how to conduct a simulation-based power analysis for non-nested model comparisons. For the sake of illustration, we will refer to the example on the compositional structure of memory outlined in section 1.3 throughout. For this example, we aim to derive estimates of required sample size to identify the tri-partite model as the best fitting model among the three competing models, if it indeed generated the data. In a step-by-step guide, we therefore show how to specify the competing models in R, how to set up different design conditions for the simulation, how to define the functions to generate, analyze and summarize synthetic data and how to execute the simulation. Based on two separate simulations, we show how the separability of theoretical constructs (section 3.1) and the reliability of the manifest variables (section 3.2) impacts estimates of statistical power for model comparisons by varying inter-factor correlations and factor loadings across different conditions of the simulation. Finally, we employ the same simulation-based approach to investigate Type-I error rates, that is erroneously identifying the tri-partite model as the best fitting model if the bi-partite or unitary model generated the data (section 3.3).

@@

The model is defined by a model string, which specifies three latent factors of memory (Pattern Separation [ps]; Pattern Completion, [pc]; Generalization, [gen]) which are measured through the manifest variables (tasks 1 to 13). Note that lavaan entails several user-friendly functions for fitting models that automatically include default settings for model components that are not explicitly specified (e.g. adding residual variances, covariances of exogenous latent variables). Thus, the model syntax can be kept very concise (check documentation of the respective function for full information on default settings).

@@

## 3.2	What is the Impact of the Reliability of Manifest Variables on Estimates of Statistical Power?  
In a second step, we further examined the effect of the construct reliability on statistical power by investigating the loading strength of the manifest variables. To this end we conducted a second simulation following the steps outlined above while varying the sample size from 50 to 200 and the loadings of the manifest variables from 0.5 to 0.9. Figure 3B displays the obtained power curves, illustrating statistical power dependent on total sample size and loading strength of the manifest variables. For this estimate, the factor inter-correlation was set to 0.3. Again, increasing sample size lead to higher statistical power. However, the loading strength impacted estimates of statistical power: the higher the loadings of the manifest variables, the higher the statistical power achieved by a given sample size. For reasonably high levels of factor loadings (>= 0.7) a statistical power of 95% was reached for *N* = 125.

@@

First, we showed that detecting differences in model fit requires increasingly many participants the higher the inter-correlation of the factors turns out practically. Therefore, it is warranted to consider the actual separability of the constructs in question to avoid underpowered sampling plans. We acknowledge that committing to specific parameter values, such as factor inter-correlations, can be difficult when setting up the population models for the simulations. Thus, we recommend researchers to consult existing literature in order to make informed assumptions on the relationship among different latent factors. For cases in which no or little evidence exists on the relationship between the constructs under investigation, we recommend to enter these inter-correlations into the simulation as parameter of interest (as demonstrated in the example of this paper) and to then carefully evaluate what level of resolution is required when separating inter-related factors depending on practical relevance.  
Second, the results on the loadings of manifest variables stresses the importance of employing indicators that capture the latent construct with high reliability [@zuo_2019]. Considering the issue of reliability is a crucial step in experimental design, as the reliability of the indicators heavily influences the practical ability to identify the targeted latent construct with sufficient precision and differentiate between competing models. Again, committing to population values can be difficult, especially if there is no prior work from which to draw values. In this case, we still advice to estimate plausible values from related scientific areas. For instance, selecting factor loading strength in the tri-partite model is not straightforward. However, we can do a literature search on similar models and take the reported test reliability (e.g., a test-retest) or construct reliability (e.g., coefficient alpha) and translate this back into factor loadings strength and residual loadings [see @brandmaier_2018 for examples on how to parametrize longitudinal models from minimal available information]. If information about test reliability is available, we set the loadings to the square-root of the reliability to take this prior information into account. Then, this yields the following construct reliability for our design (assuming identical loadings across *k* items):

@@

Note that in this tutorial, we limited our analyses to investigating the impact of either the separability of the theoretical constructs or the impact of the reliability of the estimates by keeping the other parameter constant throughout the simulation. The interaction of these parameters may differentially impact the resultant sample size estimates. However, a thorough investigation of the interaction of the two influencing factors would exponentially increase the number of conditions in a simulation. In our example on memory component processes, we have no theory-driven predictions with regards to the separability of the constructs and the reliability of the indicators’ interaction. Thus, we confine the current analysis to separate analyses in the interest of conciseness. For cases in which researchers are interested in testing a particular interaction effect between such parameters, we recommend to expand the conditions of the simulation accordingly.

While we demonstrate in this tutorial that simulation-based techniques help overcome methodological shortcomings from traditional analytical approaches, the randomness that is inherent to such approaches may seem at odds with efforts in results reproducibility and replicability. Here, a clear-cut solution is through transparent and accessible documentation of the simulation code and adequate software management. That is, researchers should make use of repositories on platforms such as GitHub or the Open Science Framework to make code scripts available to other researchers, and clearly specify the versions of software programs and packages that were used (e.g. via containerization or at least in written form). We thus urge researchers to seriously consider the issue of reproducibility for their simulation analyses, to make outcomes of such a-priori power estimates easily accessible to other researchers and reviewers [for a more detailed guideline on a reproducible workflow in R see @peikert_2021 and the implementation of this manuscript on GitHub as an example]. Note, however, that—specifically in cases when simulations are parallelized on computing clusters—the exact reproducibility of simulations as demonstrated in this tutorial might be compromised, even when a seed is specified in the analysis script. Nevertheless, a sufficiently high number of iterations ensures replicability, i.e., comparable estimates of statistical power. Derived conclusions about required sample size will therefore remain unaffected.

@@

# Open Practices Statement {-}

  The supplementary materials for this article (including the data and R scripts) are publically available at <https://github.com/ebuchberger/Estimating-Power-for-SEM>. The analyses reported in this manuscript were not preregistered.
